Data Exploration: IUCN and AquaMaps

### 2015-09-04 data_explore.Rmd
Tried a few things... not looking super for area correlation...

spp_list = spp_list%>%
              mutate(total_area=NA,
                     am_area = NA,
                     iucn_area=NA,
                     perc = NA,
                     am_perc=NA,
                     iucn_perc=NA,
                     sm_perc = NA, #filled with % of smaller range that is within the larger range
                     sm_range = NA) #what range is smaller

@@

  spp <- spp_list[i,]$sciname #species name

@@

  df <- gather(spp_list,perc.type,percent, perc,am_perc,iucn_perc)%>%
          dplyr::select(sciname,spp_group,perc.type,percent)%>%
            unique()

  perc_c = df%>%
            filter(sciname==spp)

  p1 = perc_c%>%dplyr::select(percent)
  paste0(round(p1[1,],2),'% of the total area in this map is covered by both IUCN and AquaMaps')
  paste0(round(p1[2,],2),'% of the total area in this map is covered just by AquaMaps')
  paste0(round(p1[3,],2),'% of the total area in this map is covered just by IUCN')
```

@@

all<-ggplot(df,aes(percent,fill=perc.type))+geom_histogram(binwidth=1,alpha=.3)+scale_fill_discrete(name="Legend",labels=c("Overlap", "AquaMaps only", "IUCN only"))+labs(x='Percent of total range', y='Number of Species')

@@

sp <- ggplot(df,aes(percent,fill=perc.type))+geom_histogram(binwidth=1,alpha=.3) +facet_wrap(~spp_group) +
  scale_fill_discrete(name="Legend",labels=c("Overlap", "AquaMaps only", "IUCN only"))+labs(x='Percent of total range', y='Number of Species')

@@

```{r,echo=F}

d <- spp_list%>%
  dplyr::select(sciname,sm_perc,sm_range)%>%
      unique()

ggplot(d,aes(sm_perc,fill=sm_range))+geom_histogram(binwidth=1,alpha=.7)+
  scale_fill_discrete(name="Dataset with smaller range",labels=c("AquaMaps", "IUCN"))+labs(x='Percent of smalle range that falls within the larger range', y='Number of Species')

reg <- ggplot(spp_list, aes(sm_perc,total_area,color=sm_range)) + geom_point() + geom_abline() + labs(y='Total Area', x='Percent overlap smaller range in larger', title='Comparing differences in area across species groups', fill='IUCN Category')

@@

model = lm(spp_list$total_area~spp_list$sm_perc)
summary(model)

@@

>>>>

spp_list = spp_list %>%
              mutate(total_area = NA,
                     am_area    = NA,
                     iucn_area  = NA,
                     perc       = NA,
                     am_perc    = NA,
                     iucn_perc  = NA,
                     sm_perc    = NA, #filled with % of smaller range that is within the larger range
                     sm_range   = NA) #what range is smaller

@@

  spp     <- spp_list[i, ]$sciname #species name

@@

  df <- gather(spp_list, perc.type, percent, perc, am_perc, iucn_perc) %>%
          dplyr::select(sciname, spp_group, perc.type, percent) %>%
            unique()

  perc_c = df %>%
            filter(sciname == spp)

  p1 = perc_c %>% dplyr::select(percent)
  paste0(round(p1[1, ], 2), '% of the total area in this map is covered by both IUCN and AquaMaps')
  paste0(round(p1[2, ], 2), '% of the total area in this map is covered just by AquaMaps')
  paste0(round(p1[3, ], 2), '% of the total area in this map is covered just by IUCN')
```

@@

all<-ggplot(df,aes(percent,fill=perc.type))+
    geom_histogram(binwidth=1,alpha=.3)+
    scale_fill_discrete(name="Legend",labels=c("Overlap", "AquaMaps only", "IUCN only"))+
    labs(x='Percent of total range', y='Number of Species')

@@

sp <- ggplot(df,aes(percent,fill=perc.type))+geom_histogram(binwidth=1,alpha=.3) +
  facet_wrap(~spp_group) +
  scale_fill_discrete(name="Legend",labels=c("Overlap", "AquaMaps only", "IUCN only"))+
  labs(x='Percent of total range', y='Number of Species')

@@

```{r pct small within large, echo=F}

d <- spp_list %>%
  dplyr::select(sciname, sm_perc, sm_range) %>%
      unique()

ggplot(d, aes(sm_perc, fill = sm_range)) + 
  geom_histogram(binwidth = 1, alpha = .7) +
  scale_fill_discrete(name = "Dataset with smaller range", labels = c("AquaMaps", "IUCN")) + 
  labs(x = 'Percent of smaller range that falls within the larger range', y = 'Number of Species')

``` {r pct overlap vs area, echo = FALSE}

reg <- ggplot(spp_list, aes(total_area, sm_perc, color = sm_range)) + 
  geom_point() + 
  geom_abline() + 
  labs(x = 'Total Area (km2)', y = 'Percent overlap smaller range in larger', 
       title = 'Size of species range vs. percent overlap of \nsmaller range with larger')

@@

# remove all area greater than 100000000

l <- spp_list %>% filter(total_area < 100000000)

r <- ggplot(l, aes(total_area, sm_perc, color = sm_range)) + 
  geom_point() + 
  labs(x = 'Total Area (km2)', y = 'Percent overlap smaller range in larger', 
       title = 'Size of species range vs. percent overlap of \nsmaller range with larger')

print(r)

# Log transform area

logA_vs_pct <- ggplot(spp_list, aes(log_total_area, sm_perc, color = sm_range)) + 
  geom_point() + 
  labs(x = 'Total Area (km2)', y = 'Percent overlap smaller range in larger', 
       title = 'Size of species range vs. percent overlap of \nsmaller range with larger')

print(logA_vs_pct)
```

``` {r examine regressions, echo = FALSE}

A_mdl      <- lm(total_area ~ sm_perc, spp_list)
#                Estimate Std. Error t value Pr(>|t|)    
#   (Intercept) -16285410    2858388  -5.697  1.4e-08 ***
#   sm_perc        533370      41702  12.790  < 2e-16 ***
#   Multiple R-squared:  0.07732,	Adjusted R-squared:  0.07685

A_mdl_clip <- lm(total_area ~ sm_perc, l)
#               Estimate Std. Error t value Pr(>|t|)    
#   (Intercept)  2993939     810560   3.694 0.000227 ***
#   sm_perc       153523      11952  12.844  < 2e-16 ***
#   Multiple R-squared:  0.07955,	Adjusted R-squared:  0.07906 
logA_mdl   <- lm(log_total_area ~ sm_perc, spp_list)
#                Estimate Std. Error t value Pr(>|t|)    
#   (Intercept) 14.775325   0.096446   153.2   <2e-16 ***
#   sm_perc      0.016885   0.001407    12.0   <2e-16 ***
#   Multiple R-squared:  0.0687,	Adjusted R-squared:  0.06823 
```

``` {r examine quartiles, echo = FALSE}

summary(spp_list$total_area)
#      Min.   1st Qu.    Median      Mean   3rd Qu.      Max.
#     55760   3032000  10790000  18500000  22670000 372100000

qtiles <- c(quantile(spp_list$total_area, 0.0),
            quantile(spp_list$total_area, .25),
            quantile(spp_list$total_area, .50),
            quantile(spp_list$total_area, .75),
            quantile(spp_list$total_area, 1.0))
qtile_mdl <- list()
for (i in 1:4) {
  area_clip <- spp_list %>%
    filter(total_area >= qtiles[i] & total_area < qtiles[i + 1])
  qtile_mdl[[i]] <- lm(total_area ~ sm_perc, area_clip)

  qtile_plot <- ggplot(area_clip, aes(total_area, sm_perc, color = sm_range)) + 
  geom_point() + 
  labs(x = 'Total Area (km2)', y = 'Percent overlap smaller range in larger', 
       title = 'Size of species range vs. percent overlap of \nsmaller range with larger')

print(qtile_plot)
}
# First quartile regression:
#                 Estimate Std. Error t value Pr(>|t|)    
#     (Intercept)  1423917     101053  14.091   <2e-16 ***
#     sm_perc         1188       1555   0.764    0.445    
#     Multiple R-squared:  0.001198,	Adjusted R-squared:  -0.0008529 
#
# second quartile:
#                 Estimate Std. Error t value Pr(>|t|)    
#     (Intercept)  7011028     275943  25.408  < 2e-16 ***
#     sm_perc       -17725       4228  -4.192 3.28e-05 ***
#     Multiple R-squared:  0.03483,	Adjusted R-squared:  0.03285 
#
# third quartile:
#                 Estimate Std. Error t value Pr(>|t|)    
#     (Intercept) 12813279     507856  25.230  < 2e-16 ***
#     sm_perc        64710       8159   7.931  1.5e-14 ***
#     Multiple R-squared:  0.1146,	Adjusted R-squared:  0.1128 
#
# fourth quartile:
#                   Estimate Std. Error t value Pr(>|t|)    
#     (Intercept) -101736351   17259544  -5.894 7.04e-09 ***
#     sm_perc        1911286     215216   8.881  < 2e-16 ***
#     Multiple R-squared:  0.1399,	Adjusted R-squared:  0.1381

# This avenue of exploration looks like a bust.

@@

``` {r, echo = FALSE}

# Regression

lm_eqn <- function(df){
    m  <- lm(total_area ~ sm_perc, spp_list)
    eq <- substitute(italic('Total Area') == a + b %.% italic('Percent')*","~~italic(r)^2~"="~r2, 
          list(a  = format(coef(m)[1], digits = 2), 
               b  = format(coef(m)[2], digits = 2), 
               r2 = format(summary(m)$r.squared, digits = 3)))
    as.character(as.expression(eq))             
}

p1 <- reg + geom_text(x = 50, y = 80000000, label = lm_eqn(spp_list), parse = TRUE)
